Regression models for Cylindrical data in Psychology

### 2019-01-28 Manuscript/ManuscriptMarkdownCV.tex
Update manuscript

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Regression models for Cylindrical data in Psychology},
            pdfkeywords={bla vla},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Regression models for Cylindrical data in Psychology}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}

\usepackage{multirow}
\usepackage{appendix}
\usepackage{color}
\usepackage{hyperref}
\usepackage{subcaption}
\setlength\parindent{24pt}
\usepackage[nodisplayskipstretch]{setspace}\doublespacing
\DeclareRobustCommand{\VANDER}[3]{#2} % set up for citation (in .tex %  place \DeclareRobustCommand{\VANDER}[3]{#3} before bibliography)

\begin{document}
\maketitle
\begin{abstract}
Cylindrical data are multivariate data which consist of a directional,
in this paper circular, and a linear component. Examples of cylindrical
data in psychology include human navigation (direction and distance of
movement), eye-tracking research (direction and length of saccades) and
data from an interpersonal circumplex (type and strength of
interpersonal behavior). In this paper we adapt four models for
cylindrical data to include a regression of the circular and linear
component onto a set of covariates. Subsequently, we illustrate how to
fit these models and interpret their results on a dataset on the
interpersonal behavior of teachers.
\end{abstract}

@@

Cylindrical data are data that consist of a linear variable and a
directional variable. In this paper, the directional variable is
circular, meaning that it consists of a single angle instead of a set of
angles. A circular variable is different from a linear variable in the
sense that it is measured on a different scale. Figure \ref{circline}
shows the difference between a circular scale (right) and a linear scale
(left). The most important difference is that on a circular scale the
datapoints 0\(^\circ\) and 360\(^\circ\) are connected and in fact
represent the same number while on a linear scale the two ends,
\(-\infty\) and \(\infty\), are not connected and consequently the
values 0\(^\circ\) and 360\(^\circ\) are located on different places on
the scale. This difference requires us to use special statistical
methods for circular variables (see e.g. Fisher (1995) for an
introduction to circular data and Mardia \& Jupp (2000), Jammalamadaka
\& Sengupta (2001) and Ley \& Verdebout (2017) for a more elaborate
overview). For instance, a notion as simple as the sample average needs
to be re-defined. As is the case for circular data, the analysis of
cylindrical data also requires special methods.\newline
\indent Cylindrical data occur in several fields of research, such as
for instance in meteorology (García-Portugués, Crujeiras, \&
González-Manteiga, 2013), ecology (García-Portugués, Barros, Crujeiras,
González-Manteiga, \& Pereira, 2014) or marine research (Lagona, Picone,
Maruotti, \& Cosoli, 2015). However, also in psychology several types of
cylindrical data can be found. For example, in research on human
navigation in the field of cognitive psychology both the distance, a
linear variable, and the direction, a circular variable, of movement are
of interest (Chrastil \& Warren, 2017). In eye-tracking research,
saccade data are an example of cylindrical data, because both the
direction (\emph{i.e.}, the circular variable) and the duration
(\emph{i.e.}, the linear variable) of the saccades are of interest (for
a review of eye-tracking research see Rayner (2009)).\newline
\indent The type of data that is used in the present study are also
psychological, namely data from circumplex measurement instruments. For
instance, data from the interpersonal circumplex as used in personality
psychology are by definition of a cylindrical nature (see Section
\ref{Example} for a more detailed explanation).\newline
\indent In the literature, several methods have been put forward to
model the relation between the linear and circular component of a
cylindrical variable. Some of these are based on regressing the linear
component onto the circular component using the following type of
relation: \[y = \beta_0 +
\beta_1*\cos(\theta) + \beta_2*\sin(\theta)+ \epsilon,\] where \(y\) is
the linear component and \(\theta\) the circular component (Johnson \&
Wehrly, 1978; Mardia \& Sutton, 1978; Mastrantonio, Maruotti, \&
Jona-Lasinio, 2015). Others model the relation in a different way,
e.g.~by specifying a multivariate model for several linear and circular
variables and modelling their covariance matrix (Mastrantonio, 2018) or
by proposing a joint cylindrical distribution. For example, Abe \& Ley
(2017) introduce a cylindrical distribution based on a Weibull
distribution for the linear component and a sine-skewed von Mises
distribution for the circular component and link these through their
respective shape and concentration parameters. However, none of the
methods that have been proposed thus far include additional covariates
onto which both the circular and linear component are regressed.\newline
\indent Our aim in this paper is to fill this gap in the literature by
adapting four existing cylindrical models in such a way that they
include a regression of both the linear and circular component of a
cylindrical variable onto a set of covariates. From now on we will
therefore refer to the components of the cylindrical variable as outcome
components. Additionally, we will show how a correct statistical
treatment of such cylindrical data can lead to new insights. We will do
this for the teacher data, a dataset from the field of educational
psychology. In the teacher data, apart from modelling the relation
between the linear and circular component of a cylindrical variable we
would also like to predict the two components from a set of covariates
in a regression model.

The paper is organized as follows: Section \ref{Example} describes the
teacher data, while Section \ref{Models} presents the four cylindrical
models and our associated adaptations to new regression models. In that
same section we also discuss the model fit criterion that we will use in
Section \ref{DataAnalysis} for the comparison of the four models. A
detailed data analysis with interesting new insights is also provided in
Section \ref{DataAnalysis}. We conclude the paper with a discussion in
Section \ref{Discussion}, and the Supplementary Material collects the
technical details of the MCMC procedures.

@@

The motivating example for this article comes from the field of
educational psychology and was collected for the studies on classroom
climate of \VANDER{Want}{Van der}{van der} Want (2015), Claessens (2016)
and Pennings et al. (2018). An indicator of the quality of the classroom
climate is the students' perception of their teachers' interpersonal
behavior. These interpersonal perceptions, both in educational
psychology as well as in other areas of psychology, can be measured
using circumplex measurement instruments (see Horowitz \& Strack (2011)
for an overview of many such instruments).\newline
\indent The circumplex data used in this paper are measured using the
Questionnaire on Teacher Interaction (QTI) (Wubbels, Brekelmans, Brok,
\& Tartwijk, 2006) which is one such circumplex measurement instrument.
The QTI is designed to measure student perceptions of their teachers'
interpersonal behavior and contains items that load on two interpersonal
dimensions: Agency and Communion. Agency refers to the degree of power
or control a teacher exerts in interaction with his/her students.
Communion refers to the degree of friendliness or affiliation a teacher
conveys in interaction with his/her students. The loadings on the two
dimensions of the QTI can be placed in a two-dimensional space formed by
Agency (vertical) and Communion (horizontal), see Figure \ref{QTI}.
Different parts of this space are characterized by different teacher
behavior, e.g. `helpful' or `uncertain'. This two-dimensional space is
called the interpersonal circle/circumplex (IPC). The IPC is ``a
continuous order with no beginning or end'' (Gurtman, 2009, p. 2). We
call such ordering a circumplex ordering and the IPC is therefore often
called the interpersonal circumplex. The ordering also implies that
scores on the IPC could be viewed as a circular variable.\newline
\indent Cremers, Mainhard, \& Klugkist (2018a) explain the circular
nature of the IPC data and analyze them as such using a circular
regression model. The two-dimension scores Agency and Communion can be
converted to a circular score using the two-argument arctangent function
in \eqref{PredVal}, where \(A\) represents a score on the Agency
dimension and \(C\) represents a score on the Communion dimension

\begin{equation}\label{PredVal}
\theta          = \text{atan2}\left(A, \: C\right)  =
\left\{{\begin{array}{lcl}
                                                                       \arctan\left(\frac{A}{C}\right) & \text{if}  \quad&C > 0 \\
\arctan\left(\frac{A}{C}\right) + \pi & \text{if}  \quad& C  <  0  \:\: \&\:\: A \geq 0\\
 \arctan\left(\frac{A}{C}\right) - \pi & \text{if}  \quad&C  <  0 \:\:  \&\:\:A  < 0\\
 +\frac{\pi}{2} & \text{if}  \quad& C  =  0  \:\: \&\:\:A > 0\\
 -\frac{\pi}{2} & \text{if}  \quad& C =  0  \:\: \&\:\:A < 0\\
 \text{undefined} & \text{if} \quad& C =  0   \:\: \&\:\:A = 0.
 \end{array}}
\right.
\end{equation}

The resulting circular variable \(\theta\) can then be modelled and
takes values in the interval \([0, 2\pi)\). However, when
two-dimensional data are converted to the circle we lose some
information, namely the length of the two-dimensional vector
\((A, C)^t\), \emph{i.e.}, its Euclidean norm
\(\mid\mid (A, C)^t \mid\mid\). This length represents the strength of
the type of interpersonal behavior a teacher shows towards his/her
students and can be considered as the linear variable in a cylindrical
model, allowing us to model a circular variable \(\theta\) together with
the linear variable corresponding to \(\mid\mid (A, C)^t \mid\mid\).
This leads to an improved analysis of interpersonal circumplex data as
we take all information into account. In the next section we introduce
several models that can be used for a more accurate and informative
regression analysis on the teacher data. First however we will provide
descriptives for our data set.

\subsection{Data description}\label{DataDescriptives}

The teacher data was collected between 2010 and 2015 and contains
several repeated measures on the IPC of 161 teachers. Measurements were
obtained using the QTI and taken in different years and classes. For
this paper we only consider one measurement, the first occasion (2010)
and largest class if data for multiple classes were available. In
addition to the score on the IPC, the circular outcome, and the strength
of the score on the IPC, the linear outcome, a teachers' self-efficacy
(\verb|SE|) concerning classroom management is used as covariate in the
analysis. After listwise deletion of missings (\(3\) in total, only for
the self-efficacy) we have a sample of 148 teachers. Table
\ref{Tableteacherdescriptives} shows descriptives for the dataset. Here
\(\hat{\rho}\) is a sample estimate for the circular concentration where
a value of 0 means that the data is not concentrated at all, \emph{i.e.}
spread over the entire circle, and a value of 1 means that all data is
concentrated at a single point on the circle. Figure \ref{dataplot} is a
scatterplot showing the relation between the linear and circular outcome
of the teacher data.

@@

\begin{table}[h]
\centering
\caption{Descriptives for the teacher dataset.} 
\begin{tabular}{lrrrl}
  \noalign{\smallskip}\hline\noalign{\smallskip}
Variable & mean/$\bar{\theta}$ & sd/$\hat{\rho}$ & Range & Type \\ \hline\noalign{\smallskip}
IPC &33.22$^\circ$& 0.76 & - & Circular\\
strength IPC & 0.43 & 0.15 & 0.08 - 0.80 & Linear\\
SE & 5.04 & 1.00 & 1.5 - 7.0 & Linear\\
   \hline
\end{tabular}
\label{Tableteacherdescriptives}
\end{table}

\section{Four cylindrical regression models}\label{Models}

In this section we present four cylindrical models and adapt them such
that they contain predictors for the linear and circular outcomes, \(Y\)
and \(\Theta\). The first two models are based on a construction by
Mastrantonio et al. (2015), while the other models are extensions of the
models from Abe \& Ley (2017) and Mastrantonio (2018).

\subsection{The modified CL-PN and modified CL-GPN  models}\label{CL-(G)PN}

Following Mastrantonio et al. (2015) we consider in this section two
models where the relation between \(\Theta \in [0, 2\pi)\) and
\(Y\in (-\infty, + \infty)\) and \(q\) covariates is specified as

\begin{equation}\label{circlinlink}
Y = \gamma_0 + \gamma_{cos}*\cos(\Theta)*R + \gamma_{sin}*\sin(\Theta)*R + \gamma_1*x_1 + \dots + \gamma_q*x_q +  \epsilon,
\end{equation}

\noindent where the random variable \(R\geq0\) will be introduced below,
the error term \(\epsilon \sim N(0, \sigma^2)\) with variance
\(\sigma^2>0\),
\(\gamma_0, \gamma_{cos}, \gamma_{sin}, \gamma_1, \dots, \gamma_q\) are
the intercept and regression coefficients and \(x_1, \dots, x_q\) are
the \(q\) covariates. In both of these models the conditional
distribution of \(Y\) given \(\Theta=\theta\) and \(R = r\) is given by

\begin{equation}\label{ycondtheta}
f(y \mid \theta, r) = \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left[-\frac{(y - (\gamma_0 + \gamma_1x_1 + \dots + \gamma_qx_q+c))^{2}}{2\sigma^2}\right],\nonumber
\end{equation}

>>>>

\documentclass[man]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Regression models for Cylindrical data in Psychology},
            pdfauthor={Jolien Cremers, Helena J.M. Pennings, \& Christophe Ley},
            pdfkeywords={cylindrical data, regression, interpersonal behavior},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Regression models for Cylindrical data in Psychology}
    \author{Jolien Cremers\textsuperscript{1}, Helena J.M.
Pennings\textsuperscript{2,3}, \& Christophe Ley\textsuperscript{4}}
    \date{}

\shorttitle{Cylindrical data in Psychology}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Department of Methodology and Statistics, Utrecht University\\\textsuperscript{2} TNO\\\textsuperscript{3} Department of Education, Utrecht University\\\textsuperscript{4} Department of Applied Mathematics, Computer Science and Statistics, Ghent University}
\keywords{cylindrical data, regression, interpersonal behavior\newline\indent Word count: X}
\usepackage{csquotes}
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

\usepackage{longtable}
\usepackage{lscape}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage[flushleft]{threeparttable}
\usepackage{threeparttablex}

\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}

\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

\usepackage{lineno}

\linenumbers
\DeclareRobustCommand{\VANDER}[3]{#2}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{color}
\usepackage{subcaption}

\authornote{Add complete departmental affiliations for each
author here. Each new line herein must be indented, like this line.

Enter author note here.

Correspondence concerning this article should be addressed to Jolien
Cremers, Postal address. E-mail:
\href{mailto:joliencremers@gmail.com}{\nolinkurl{joliencremers@gmail.com}}}

\abstract{
Cylindrical data are multivariate data which consist of a directional,
in this paper circular, and a linear component. Examples of cylindrical
data in psychology include human navigation (direction and distance of
movement), eye-tracking research (direction and length of saccades) and
data from an interpersonal circumplex (type and strength of
interpersonal behavior). In this paper we adapt four models for
cylindrical data to include a regression of the circular and linear
component onto a set of covariates. Subsequently, we illustrate how to
fit these models and interpret their results on a dataset on the
interpersonal behavior of teachers.

}

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}
\maketitle

@@

In social sciences the use of cylindrical data is very common. Such data
consist of a linear and a circular component. Gurtman (2011) refers to
such data as vectors, with a directional measure (i.e., the circular
component) and a measure indicating the magnitude (i.e., linear
component). Many established models in psychology are often referred to
as circular or circumplex models, but those models are cylindrical.
Examples of such cylindrical models are the interpersonal
circle/circumplex (Leary, 1957; Wiggins, 1996; Wubbels, Brekelmans,
Brok, \& Tartwijk, 2006), the circumplex of affect (Russell, 1980), the
circumplex of human emotion (Plutchik, 1997) or the model of human
values (Schwartz, 1992). \newline \indent Also, many of the more recent
types of data that are studied in psychology are cylindrical. For
example, research on human navigation uses data where distance (i.e.,
linear component) and direction (i.e., the linear component) are of
interest (Chrastil \& Warren, 2017) or in eye-tracking, the saccade data
also consist of both the direction (i.e., the circular variable) and the
duration (i.e., the linear variable)(e.g., Rayner (2009)). Apart from
the social sciences data with a circular and linear outcome more
commonly occur in meteorology meteorology (García-Portugués, Crujeiras,
\& González-Manteiga, 2013), ecology (García-Portugués, Barros,
Crujeiras, González-Manteiga, \& Pereira, 2014) or marine research
(Lagona, Picone, Maruotti, \& Cosoli, 2015) \newline \indent Up until
now researchers studying cylindrical data had to rely on the linear
statistical methods to analyze their research results. Focusing on only
the circular variable or only on the linear variable. However, lately
more and more of these researchers acknowledge that linear methods are
not sufficient and call for new methods (Gurtman, 2011; Pennings, 2017b;
Wright, Pincus, Conroy, \& Hilsenroth, 2009) that take into account both
the circular and the linear component of these data. \newline
\indent The aim in the present paper is twofold. Firstly, we intend to
fill this gap in the literature on cylindrical models by adapting four
existing cylindrical models in such a way that they include a regression
of both the linear and circular component of a cylindrical variable onto
a set of covariates. From now on we will therefore refer to the
components of the cylindrical variable as outcome components. Secondly,
we will show how using these models can benefit the analysis of
circumplex data and cylindrical data in psychology in general. More
specifically we will show these benefits for interpersonal teacher data
from the field of educational psychology. Apart from modelling the
dependence between the linear and circular component of a cylindrical
variable we would also like to predict the two components from a set of
covariates in a regression model. In our analysis, we will show how this
can be done in four different ways using the adapted cylindrical models.

\subsection{Cylindrical data}

Thus, data that consist of a variable and a circular variable is called
cylindrical data. A circular variable is different from the linear
variable in the sense that it is measured on a different scale. Figure
\ref{circline} shows the difference between a circular scale (right) and
a linear scale (left). The most important difference is that on a
circular scale the datapoints 0\(^\circ\) and 360\(^\circ\) are
connected and in fact represent the same number while on a linear scale
the two ends, \(-\infty\) and \(\infty\), are not connected and
consequently the values 0\(^\circ\) and 360\(^\circ\) are located on
different places on the scale. Both circular data and cylindrical data
require special analysis methods due to this periodicity in the scale of
a circular variable (see e.g.~Fisher (1995) for an introduction to
circular data and Mardia and Jupp (2000), Jammalamadaka and Sengupta
(2001) and Ley and Verdebout (2017) for a more elaborate
overview).\newline \indent In the literature, several methods have been
put forward to model the relation between the linear and circular
component of a cylindrical variable. Some of these are based on
regressing the linear component onto the circular component using the
following type of relation: \[y = \beta_0 +
\beta_1*\cos(\theta) + \beta_2*\sin(\theta)+ \epsilon,\] where \(y\) is
the linear component and \(\theta\) the circular component (Johnson \&
Wehrly, 1978; Mardia \& Sutton, 1978; Mastrantonio, Maruotti, \&
Jona-Lasinio, 2015). Others model the relation in a different way,
e.g.~by specifying a multivariate model for several linear and circular
variables and modelling their covariance matrix (Mastrantonio, 2018) or
by proposing a joint cylindrical distribution. For example, Abe and Ley
(2017) introduce a cylindrical distribution based on a Weibull
distribution for the linear component and a sine-skewed von Mises
distribution for the circular component and link these through their
respective shape and concentration parameters. However, none of the
methods that have been proposed thus far include additional covariates
onto which both the circular and linear component are regressed.

@@

The motivating example for this article comes from the field of
educational psychology and was collected for the studies on classroom
climate of \VANDER{Want}{Van der}{van der} Want (2015), Claessens (2016)
and Pennings (2017a). An indicator of the quality of the classroom
climate is the students' perception of their teachers' interpersonal
behavior. These interpersonal perceptions, both in educational
psychology as well as in other areas of psychology, can be measured
using circumplex measurement instruments (see Horowitz and Strack (2011)
for an overview of many such instruments).\newline \indent The
circumplex data used in this paper are measured using the Questionnaire
on Teacher Interaction (QTI) (Wubbels et al., 2006) which is one such
circumplex measurement instrument. The QTI is designed to measure
student perceptions of their teachers' interpersonal behavior and
contains items that load on two interpersonal dimensions: Agency and
Communion. Agency refers to the degree of power or control a teacher
exerts in interaction with his/her students. Communion refers to the
degree of friendliness or affiliation a teacher conveys in interaction
with his/her students. The loadings on the two dimensions of the QTI can
be placed in a two-dimensional space formed by Agency (vertical) and
Communion (horizontal), see Figure \ref{QTI}. Different parts of this
space are characterized by different teacher behavior,
e.g.~\enquote{helpful} or \enquote{uncertain}. This two-dimensional
space is called the interpersonal circle/circumplex (IPC). The IPC is
``a continuous order with no beginning or end'' (Gurtman, 2009, p. 2).
We call such ordering a circumplex ordering and the IPC is therefore
often called the interpersonal circumplex. The ordering also implies
that scores on the IPC could be viewed as a circular variable. This
circular variable represents the type of interpersonal behavior that a
teacher shows towards his/her students.\newline \indent Cremers et al.
(2018a) explain the circular nature of the IPC data and analyze them as
such using a circular regression model. The two dimension scores, Agency
and Communion, can be converted to a circular score using the
two-argument arctangent function in
\eqref{PredVal}\footnote{The selection of
the origin in circumplex data depends on the scaling of the Agency and Communion
scores; their respective 0 scores form the origin. Agency and Communion are
measured on a scale from 1 to 5 and for analysis purposes they are later
converted to scale ranging from -1 to 1. Scaling is however only considered an
issue in those instances where cylindrical data is derived from measurements in
bivariate space.}, where \(A\) represents a score on the Agency
dimension and \(C\) represents a score on the Communion dimension. Note
that when placing a unit circle on Figure \ref{QTI} we see that the
Agency dimension is related to the sine of the circular score and the
Communion dimension is related to the cosine of the circular score.
\begin{equation}\label{PredVal}
\theta          = \text{atan2}\left(A, \: C\right)  =
\left\{{\begin{array}{lcl}
                                                                       \arctan\left(\frac{A}{C}\right) & \text{if}  \quad&C > 0 \\
\arctan\left(\frac{A}{C}\right) + \pi & \text{if}  \quad& C  <  0  \:\: \&\:\: A \geq 0\\
 \arctan\left(\frac{A}{C}\right) - \pi & \text{if}  \quad&C  <  0 \:\:  \&\:\:A  < 0\\
 +\frac{\pi}{2} & \text{if}  \quad& C  =  0  \:\: \&\:\:A > 0\\
 -\frac{\pi}{2} & \text{if}  \quad& C =  0  \:\: \&\:\:A < 0\\
 \text{undefined} & \text{if} \quad& C =  0   \:\: \&\:\:A = 0.
 \end{array}}
\right.
\end{equation} The resulting circular variable \(\theta\) can then be
modelled and takes values in the interval \([0, 2\pi)\) or
\([0^\circ, 360^\circ)\). Note that the round brackets mean that
\(2\pi\) and \(360^\circ\) are not included in the interval since these
represent the same value as 0 as a result of periodicity. \newline
\indent A circular analysis of circumplex data has several benefits: it
is more in line with its theoretical definition and it allows us to
analyse the blend of the two dimensions Agency and Communion instead of
both dimensions separately providing us with new insights compared to a
separate analysis of the two dimensions that is standard in the
literature (see e.g.~Pennings et al. (2018), Wright et al. (2009) or
Wubbels et al. (2006)). There is however one main drawback: when
two-dimensional data are converted to the circle we lose some
information, namely the length of the two-dimensional vector
\((A, C)^t\), \emph{i.e.}, its Euclidean norm
\(\mid\mid (A, C)^t \mid\mid\). This length represents the strength of
the interpersonal behavior a teacher shows towards his/her students. In
a cylindrical model this strength (the linear outcome) can be modeled
together with the type of interpersonal behavior of a teacher (the
circular outcome). This leads to an improved analysis of interpersonal
circumplex data, over either analyzing the two dimensions separately or
using a circular model, because we take all information, circular and
linear, into account. In the next section we introduce several
cylindrical models that can be used to analyze the teacher data. First
however we will provide descriptives for the teacher data.

\subsection{Data Description}\label{DataDescriptives}

The teacher data was collected between 2010 and 2015 and contains
several repeated measures on the IPC of 161 teachers. Measurements were
obtained using the QTI and taken in different years and classes. For
this paper we only consider one measurement, the first occasion (2010)
and largest class if data for multiple classes were available. This
results in a sample of 151 teachers. In addition to the type of
interpersonal behavior (IPC), the circular outcome, and the strength of
interpersonal behavior (IPC strength), the linear outcome, a teachers'
self-efficacy (\verb|SE|) concerning classroom management is used as
covariate in the analysis. In previous research, in psychology and
education it had been shown that higher self-efficacy is related to the
quality of interpersonal interactions (Locke \& Sadler, 2007;
\VANDER{Want}{Van der}{van der} Want et al., n.d.). After listwise
deletion of missings (\(3\) in total, only for the self-efficacy) we
have a sample of 148 teachers. Table \ref{Tableteacherdescriptives}
shows descriptives for the dataset. For the circular variable IPC we
show sample estimates for the circular mean \(\bar{\theta}\) and
concentration \(\hat{\rho}\). The circular concentration lies between 0,
meaning the data is not concentrated at all \emph{i.e.} spread over the
entire circle, and 1, meaning all data is concentrated at a single point
on the circle The population values of these parameters are usually, and
also in this paper, referred to as \(\mu\) (circular location) and
\(\kappa\) (circular concentration). For the linear variables (strength
IPC and SE) we show sample estimates of the linear mean and standard
deviation. Figure \ref{dataplot} is a scatterplot showing the relation
between the linear and circular outcome of the teacher data for teachers
with low SE (below 1 sd below the mean), average SE (between 1 sd below
and 1 sd above the mean) and high SE (above 1 sd above the mean).

@@

\begin{table}[h]
\centering
\caption{Descriptives for the teacher dataset.} 
\begin{tabular}{lrrrl}
  \noalign{\smallskip}\hline\noalign{\smallskip}
Variable & mean/$\bar{\theta}$ & sd/$\hat{\rho}$ & Range & Type \\ \hline\noalign{\smallskip}
IPC &33.22$^\circ$& 0.76 & - & Circular\\
strength IPC & 0.43 & 0.15 & 0.08 - 0.80 & Linear\\
SE & 5.04 & 1.00 & 1.5 - 7.0 & Linear\\
   \hline
\multicolumn{5}{l}{Note: For the circular variable IPC we show sample }\\
\multicolumn{5}{l}{estimates for the circular mean $\bar{\theta}$ and concentration $\hat{\rho}$.}\\
\multicolumn{5}{l}{For the linear variable we show the sample mean,}\\
\multicolumn{5}{l}{standard deviation and range.}
\end{tabular}
\label{Tableteacherdescriptives}
\end{table}

\section{Four Cylindrical Regression Models}\label{Models}

One of the goals of this paper is to show the benefits of cylindrical
methods for the analysis of circumplex data and cylindrical data in
psychology in general. To do so we decide to focus on four cylindrical
models. The models were selected for their relatively low complexity and
the ease with which a regression structure could be incorporated. But
also because they show different ways of modelling the the linear and
circular outcome and thereby illustrate a wider range of cylindrical
models available in the literature. All four cylindrical models contain
a set of \(q\) predictors \(\boldsymbol{x} = x_1, \dots, x_q\) and \(p\)
predictors \(\boldsymbol{z} = z_1, \dots, z_q\) for the linear and
circular outcomes, \(Y\) and \(\Theta\), respectively. The first two
models are based on a construction by Mastrantonio et al. (2015), while
the other models are extensions of the models from Abe and Ley (2017)
and Mastrantonio (2018). The four cylindrical models are introduced
separately in the subsections below. However, to provide a more succinct
overview and comparison of the four models, \ref{TableModels} gives an
overview of the similarities and differences between the models.

\begin{sidewaystable}[h]
\centering
\caption{Comparison of the four cylindrical regression models} 
\begin{tabular}{lllll}
  \noalign{\smallskip}\hline\noalign{\smallskip}
\multicolumn{1}{l}{Aspect} & CL-PN & CL-GPN  & Abe-Ley  & GPN-SSN \\ \hline\noalign{\smallskip}
$\Theta$ & &&&\\
$\:\:$Distribution& PN & GPN & Sine-skewed vM & GPN\\
$\:\:$Domain & $[0, 2\pi)$ & $[0, 2\pi)$ & $[0, 2\pi)$ & $[0, 2\pi)$\\
$\:\:$Shape & symmetric, & assymetric, & assymetric, & assymetric, \\
            & unimodal  & multimodal & unimodal   & multimodal \\\hline\noalign{\smallskip}
$Y$& &&&\\
$\:\:$Distribution & Normal & Normal & Weibull & skewed-Normal\\
$\:\:$Domain & $(-\infty, + \infty)$ & $(-\infty, + \infty)$ & $(0, + \infty)$ & $(0, + \infty)$\\
$\:\:$Shape & symmetric, & symmetric, & assymetric, & assymetric, \\
            & unimodal  & unimodal  & unimodal   & unimodal\\\hline\noalign{\smallskip}
$\Theta$-$Y$ dependence &                                   &                                   & & \\
                        & $y$ regressed on                  & $y$ regressed on                  & $\alpha$ and $\kappa$ & multivariate \\
                        & $\sin(\theta)$ and $\cos(\theta)$ & $\sin(\theta)$ and $\cos(\theta)$ & & distribution\\\hline
\multicolumn{5}{l}{Note: PN and GPN refer to the projected normal and general projected normal distribution.}\\
\multicolumn{5}{l}{vM refers to the von-Mises distribution}\\

\end{tabular}
\label{TableModels}
\end{sidewaystable}
